Widespread nitrous oxide undersaturation in U.S. lakes and reservoirs

### 2023-08-24 manuscript/manuscript_file.Rmd
manuscript updates after expaning to 2000 realizations

- incorporate in Methods section how/why size is incorporated as categorical variable and continuos variable.
-  Mike will take a stab at big SI table.
-  Citation “epa_data” isn’t populating citation 11 in bibliography.  The bibTex entry is classified as a “website”.  Roy, can you attempt to make this citation play nice with nature_geosciences.csl?

## Psuedo abstract
Lakes, ponds, and reservoirs are thought to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018], but the relative lack of published N~2~O emission rates and recent evidence of N~2~O uptake across a broad range of waterbodies[@Soued_etal_2016;@Webb_etal_2019;@Kortelainen_etal_2020] have called into question the accuracy of emission estimates.   Here we measured dissolved N~2~O concentrations and calculated emission rates across `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` **[weird, I have 1090, but NLA17 report says 1,005? I get 984 if I exclude hand picked sites.  NLA17 site info file shows 1210 sites sampled.  I think the differences between these values are related to how hand picked and state intensification sites are handled.  The sampled we used contains data from 1090 sites, so that is what we should report here.]** waterbodies in the continental U.S. (CONUS).  Nitrous oxide undersaturation was widespread through the CONUS, with `r propN2oSinkMean` % of the waterbodies functioning as sinks for atmospheric N~2~O. Despite widespread N~2~O uptake by CONUS waterbodies, a small proportion of the waterbodies in each ecoregion had sufficiently high N~2~O emission rates to offset N~2~O uptake in the majority of waterbodies, such that all ecoregions functioned as net N~2~O sources to the atmosphere.  Waterbodies with high nitrate (NO~3~) tended to be sources of N~2~O whereas those with low NO~3~ were generally sinks.  Nitrate interacted with waterbody size, however, with large waterbodies having N~2~O concentrations closer to equilibrium than small waterbodies. We estimate that CONUS waterbodies emit `r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` )metric tons N~2~O year^-1^, equivalent to only 5% of N~2~O emissions from all CONUS surface waters (e.g. zero order streams through river mouths) per IPCC estimates .  The demonstrated capacity for widespread N~2~O uptake in CONUS waterbodies highlights the need to revisit global N~2~O models which presume surface waters are a N~2~O source [@epa_2023].

## Introduction
Lakes, reservoirs, and ponds (waterbodies hereafter) are considered globally significant sources of nitrous oxide (N~2~O)[@DelSontro_etal_2018], a potent greenhouse gas and contributor to stratospheric ozone destruction.  It is generally assumed that N~2~O emissions from surface waters scale with inorganic nitrogen loading [@Mosier_etal_1998] or productivity [@DelSontro_etal_2018], but recent reports of N~2~O uptake across waterbodies in Canada [@Soued_etal_2016;@Webb_etal_2019] and Finland [@Kortelainen_etal_2020] have called this assumption into question.  Furthermore, detailed investigations of individual waterbodies have shown that lentic N~2~O biogeochemistry is dynamic, with some waterbodies alternating between N~2~O sources and sinks in response to nitrogen availability, thermal stratification, and other factors [@Beaulieu_etal_2015;@Beaulieu_etal_2014;@Musenze_etal_2014].

While recent data sets demonstrate that waterbodies can function as N~2~O sinks, these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading).  Most work on aquatic N~2~O biogeochemistry has focused on streams and rivers, but these systems have been consistently found to be N~2~O sources and the controls on N~2~O biogeochmistry likely differ between flowing and standing waters.  These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. We modeled the data using a Bayesion hierchical approach and predicted the dissolved N~2~O concentration for all 465,897 CONUS waterbodies in the target population. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`), the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`), and \>65% of the waterbodies in every ecoregion were undersaturated **[(Table 1)]**.

@@

# create flextable
flextable(table1_dat) %>%
  set_header_labels('waterbody surface area' = "total waterbody surface area\n(hectares)",
                    'N2O undersaturation extent' = "N2O undersaturation extent\n(% waterbodies)",
                    'N2O emission rate' = "N2O emission rate\n(mg N2O m-2 d-1)",
                    'N2O flux' = "N2O flux\n(metric ton N2O y-1)") %>% # 1 megagram (Mg) = 1 1000 kg = 1 metric ton
  align(j=c(2:6), align = "center", part = "all") %>% # center align columns 2:4
  valign(valign = "top", part = "header") %>% # top align column names
  # caption doesn't show in .Rmd viewer, but renders correctly in .docx
  set_caption(
    caption = as_paragraph( # allows for formatting
      as_chunk("Table 1. Mean and 95% confidence interval of N2O undersaturation extent, saturation ratio, emission rate, and flux at the national and ecoregional scales.", 
               props = fp_text_default(bold = TRUE) # here is where we specify formatting (font, size, bold, etc)
      )
    ),
    align_with_table = FALSE) %>% # table is center aligned, ignore that and apply default left align to caption

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the mean N~2~O saturation ratio (ratio of measured to equilibrium concentration) was equal to or greater than 1 in seven of the nine ecoregions **[(Table 1, Fig. 2)]**, suggesting that, on average, waterbodies in these ecoregions were supersatured with N~2~O and were therefore functioning as a source of N~2~O. The apparent contradiction between widespread N~2~O undersaturation throughout the CONUS and average dissolved N~2~O concentrations that exceed equilibrium in most ecoregions can be attributed to the distribution of dissolved N~2~O concentrations across the population of CONUS waterbodies **[(Fig. 2)]**. Across the CONUS, and within each of the nine major ecoregions, the distribution of N~2~O saturation ratios had median values that were less than 1, indicating undersaturation, but a long right tail composed of relatively rare waterbodies with levels of N~2~O supersaturation that are sufficiently high to drive the population mean to values close to, or even greater than 1. Thus most waterbodies in the CONUS are functioning as N~2~O sinks during the summer months, but a few function as strong N~2~O sources.

![Figure 2. Distribution of N~2~O saturation values for the population of waterbodies in the nine major ecoregions. Each blue line (n=500) represents the distribution of values from a single realization of the posterior prediction. Vertical black dashed and solid lines indicated median and mean, respectively. Vertical red line indicates a saturation ratio of 1.  x-axis is truncated at 3 to better show the area where 97% of the observations occur.](manuscript_figures/figure2.tiff)

@@

![Figure 3.  Conditional effects plot illustrating A) the relationship between N~2~O saturation and nitrate (NO~3~) concentration, dependent upon surface water temperature.  The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and < 4.09, 3) $\geq$ 4.09 and < 30.20, 4) $\geq$ 30.20 and < 223.13, and 5) $\geq$ 223.13, and B) relationship between N~2~O saturation and surface water temperature, dependent upon NO~3~ category.  Horizontal dashed lines in both panels reflects an N~2~O saturation value of 1.](manuscript_figures/sat_cond_surftemp.tiff)

@@

N~2~O saturation also increased with water temperature **[Fig. 3]**, which is consistent with N~2~O production and consumption occurring through temperature sensitive biological processes, but the relationship was dependent on NO~3~ availability. At low NO~3~ concentrations, temperature had a relatively minor affect on N~2~O saturation, possibly because N~2~O production/consumption was limited by substrate rather than temperature.  The effect of temperature became more pronounced when NO~3~ availability was greater, possibly reflecting the stimulatory affect of higher temperature on biological N~2~O production.  The relative effect of NO~3~ was greater than that of temperature, thus elevated water temperature serves to magnify rates of N~2~O production that are largely dictated by NO~3~ availability.

@@

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller waterbodies. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies. Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished.  While previous surveys have documented relationships between lake size and dissolved methane concentration [@Pighini_etal_2018;@Juutinen_etal_2009;@Borges_etal_2011], to the best of our knowledge this is the first study to document a relationship between lake size and $\Delta$ N~2~O.

![Figure 4.  A) Relationship between $\Delta$ N~2~O (difference between observed and equilibrium N~2~O concentration) and waterbody surface areas.  B,C) relationship between N~2~O saturation and waterbody size for the Coastal Plains and Northern Plains ecoregions.  The vertical blue line references a N~2~O saturation value of 1.0.  The horizontal red arrows highlight $\Delta$ N~2~O for the smallest lake size class.  Mean (points) and 95% credible interview (horizontal bars) are derived from 500 realizations of the poterior predictive distribution. ](manuscript_figures/deltaN2ObySize.tiff)

@@

![Figure 5. Conditional effects plots illustrating the relationship between N~2~O saturation and A) nitrate (NO~3~ availability dependent upon lake size, and B) lake size dependent upon NO~3~ availability. The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and < 4.09, 3) $\geq$ 4.09 and < 30.20, 4) $\geq$ 30.20 and < 223.13, and 5) $\geq$ 223.13.](manuscript_figures/sat_cond_area.tiff)

The model indicates that NO~3~ exerts a strong control on N~2~O, but the relationship is complicated by interactions with water temperature and lake surface area.  This finding mirrors reports from a survey of farm ponds in Canada where N~2~O was found to be partially controlled by dissolved inorganic nitrogen (DIN) concentration, but the effect of DIN was conditional upon and thermal stratification strength.  These complex interactions may explain contrasting literature reports regarding the role of NO~3~ in regulating N~2~O.  Several studies have reported positive correlations between NO~3~ and N~2~O in lakes [@Kortelainen_etal_2020;@Yang_etal_2015] while others have found no relationship [@Soued_etal_2016;@Whitfield_etal_2011].  It is possible that the NO~3~ ~ N~2~O relationship wasn't detected in the latter studies because it was confounded by interactions with temperature, lake size, thermal stratification, or other environmental variables.

Interactions between NO~3~, water temperature, and lake size produced a broad range of N~2~O saturation values across the population of CONUS waterbodies, but also enable the identifcation of broad scale patterns. First, small waterbodies with low NO~3~ concentration tend to be undersaturated, whereas small waterbodies with high NO~3~ tend to be highly supersaturated. Large waterbodies, regardless of NO~3~ status, tend to have N~2~O saturation ratios closer to equilibrium. Therefore intra and inter regional patterns in N~2~O saturation can be well predicted by the joint distributions of waterbody size, NO~3~ availability, and water temperature.

@@

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which is expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015;@Kortelainen_etal_2020;@Webb_etal_2019].  Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was positive (`r national_stats$post_med_en2o` (95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^)) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**. At the ecoregional scale, only the Northern Appalachians and Northern Plains ecoregions had mean emission rates less than 0.  These ecoregions also had the lowest mean NO~3~ availability of all nine ecoregions **[(Table S1)]**.

## Fluxes 
Total N~2~O flux from the surface of CONUS waterbodies is a function of the N~2~O emission rate and waterbody surface area.  The largest lake size class (> 50 Ha) contained only `r round(1 - proportionSmallNumber, 2) * 100`% of the total number of lakes in the population, but had `r round(1-proportionSmallSA, 2) * 100`% of total lake surface area.  This lake size class had the greatest N~2~O flux, but flux estimates for this lake size class were also the most uncertain **[(Fig. 6)]**. Small uncertainties in emission rates translate to large uncertainties in flux when expressed across the surface area of large waterbodies.  Despite widespread N~2~O undersaturation, all ecoregions were a net source of N~2~O to the atmosphere **[(Table 1)]**.  Differences in N~2~O flux across ecoregions are explained by patterns in N~2~O saturation and waterbody surface **[(Table 1)]**.  The Coastal Plains ecoregion had the greatest flux, driven by the high levels of N~2~O saturation and extensive waterbody surface area. By contrast, the Northern Plains had the lowest N~2~O saturation, waterbody surface area, and N~2~O flux.

@@

**[extract n2o data from Lauerwald for CONUS and compare]**

@@

## Conclusion
Our survey is the first to demonstrate that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer.  Despite widespread N~2~O uptake, CONUS waterbodies function as a net source of N~2~O due to high emissions from from a subset of the population.  Dissolved N~2~O in lakes and reservoirs can be predicted from non-linear interactions between water temperature, NO~3~ availability, and waterbody size.  NO~3~ availability largely determines whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink is exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increases with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates.

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these waterbodies is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved N~2~O translate to large uncertainties in flux when propagated across the surface of large waterbodies. While our sample included some of the largest waterbodies in the CONUS **[(Table S2)]**, efforts to improve the accuracy of the lentic N~2~O flux estimate should include measurements at the nations largest lakes and reservoirs.

@@

<div id="refs"></div>

@@

## Sampling
Samples were collected from approximately the middle of each waterbody for analysis of algal toxins (microcystins and cylindrospermopsin), benthic macroinvertebrates,  phytoplankton, atrazine, water chemistry, chlorophyll‐a, zooplankton, and dissolved N~2~O.  Duplicate dissolved N~2~O  samples were collected approximately 0.05 m below the water surface by drawing 105 mL of water into a 140 mL syringe equipped with a stopcock and containing 35 mL of air. The syringes were gently shaken for five minutes and the headspace gas transferred to a 30 mL syringe.  Finally, the gas sample was transferred into a 12 mL pre-evacuated (<50 mTorr) glass vial (Labco, Wales, UK) sealed with a PTFE/Silicone stacked on top of a grey chlorobutyl septa.  The temperatue of the water in the syringe was recorded immediately after the gas was transferred into the vial.  Duplicate 20 mL air samples were collected from approximately 2 m above the water surface and transferred to 12 mL glass vials.  A vertical profile of water temperatue was measured at the same site.

>>>>

-   incorporate in Methods section how/why size is incorporated as categorical variable and continuos variable.
-   Mike will take a stab at big SI table.
-   Citation "epa_data" isn't populating citation 11 in bibliography. The bibTex entry is classified as a "website". Roy, can you attempt to make this citation play nice with nature_geosciences.csl?
-   netCDF data from Lauerwald
-   Letters vs article format at Nature Geosciences
-   Add quantitative NO3 values to Fig.3B and Fig. 5B legend.
-   Add quantitative size category values to Fig. 5A legend.

## Psuedo abstract

Lakes, ponds, and reservoirs are thought to be globally important sources of nitrous oxide (N~2~O) to the atmosphere [@DelSontro_etal_2018a], but the relative lack of published N~2~O emission rates and recent evidence of N~2~O uptake across a broad range of waterbodies[@Soued_etal_2016; @Webb_etal_2019; @Kortelainen_etal_2020] have called into question the accuracy of emission estimates[@Lauerwald_etal_2019]. Here we measured dissolved N~2~O concentrations and calculated emission rates across `r dg %>% filter(sitetype == "PROB") %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` waterbodies in the continental U.S. (CONUS). Nitrous oxide undersaturation was widespread through the CONUS, with `r propN2oSinkMean` % of the waterbodies functioning as sinks for atmospheric N~2~O. Despite widespread N~2~O uptake by CONUS waterbodies, a small proportion of the waterbodies in each ecoregion had sufficiently high N~2~O emission rates to offset N~2~O uptake in the majority of waterbodies, such that all ecoregions functioned as net N~2~O sources to the atmosphere. Waterbodies with high nitrate (NO~3~) tended to be sources of N~2~O whereas those with low NO~3~ were generally sinks. Nitrate interacted with waterbody size, however, with large waterbodies having N~2~O concentrations closer to equilibrium than small waterbodies. We estimate that CONUS waterbodies emit `r national_stats %>% select(post_med_fn2o) %>% pull() %>% round(., 0)` (`r national_stats %>% select(LCL_fn2o) %>% pull() %>% round(., 0)` - `r national_stats %>% select(UCL_fn2o) %>% pull() %>% round(., 0)` ) metric tons N~2~O year^-1^, equivalent to only `r (national_stats %>% select(UCL_fn2o) %>% pull() / 22767) * 100`% of N~2~O emissions from all CONUS surface waters (e.g. zero order streams through river mouths) per IPCC estimates . The demonstrated capacity for widespread N~2~O uptake in CONUS waterbodies highlights the need to revisit global N~2~O models which presume surface waters are a N~2~O source [@epa_2023].

## Introduction

Lakes, reservoirs, and ponds (waterbodies hereafter) are considered globally significant sources of nitrous oxide (N~2~O)[@DelSontro_etal_2018a], a potent greenhouse gas and contributor to stratospheric ozone destruction. It is generally assumed that N~2~O emissions from surface waters scale with inorganic nitrogen loading [@Mosier_etal_1998] or productivity [@DelSontro_etal_2018], but recent reports of N~2~O uptake across waterbodies in Canada [@Soued_etal_2016; @Webb_etal_2019] and Finland [@Kortelainen_etal_2020] have called this assumption into question. Furthermore, detailed investigations of individual waterbodies have shown that lentic N~2~O biogeochemistry is dynamic, with some waterbodies alternating between N~2~O sources and sinks in response to nitrogen availability, thermal stratification, and other factors [@Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014].

While several measurement campaigns have demonstrated that waterbodies can function as N~2~O sinks[@Soued_etal_2016; @Webb_etal_201; @Kortelainen_etal_2020; @Beaulieu_etal_2015; @Beaulieu_etal_2014; @Musenze_etal_2014], these studies have been conducted at limited spatial scales and encompassed small ranges in potential environmental drivers (e.g. nitrogen loading). Most work on aquatic N~2~O biogeochemistry has focused on streams and rivers, but these systems have been consistently found to be N~2~O sources [@Beaulieu_etal_2008; @Beaulieu_etal_2010; @Quick_etal_2019] and the controls on N~2~O biogeochmistry likely differ between flowing and standing waters. These limitations in data availability have complicated efforts to incorporate N~2~O consumption into freshwater N~2~O budgets.

@@

In this study we present the first national-scale survey of N~2~O concentrations and emissions from CONUS waterbodies. A generalized random-tessellation stratified survey design [@Stevens_Olsen_2004] was used to select a spatially-balanced and representative sample of `r dg %>% distinct(site.id) %>% summarize(n = n()) %>% pull()` **[Roy has 1079. Handpicked and NA sites not used. Update this code.]**sites from CONUS waterbodies that are least 1 hectare in area, at least 1 meter deep, have at least 0.1 hectare of open water, and have a minimum residence time of one week **[(Fig. 1)]**. Sample sites were distributed across all nine major CONUS ecoregions and varied widely in water chemistry, morphometry, watershed land-use, and climate [@epa_data]. `r undersatN2oPercent` percent of the sampled waterbodies were undersaturated in N~2~O and were therefore functioning as N~2~O sinks **[(Fig. 1)]**. We modeled the data using a Bayesian hierarchical approach and predicted the dissolved N~2~O concentration for all 465,897 CONUS waterbodies in the target population. At the population level, `r propN2oSinkMean` percent (95% CI: `r propN2oSinkLower` - `r propN2oSinkUpper`) of CONUS waterbodies were functioning as N~2~O sinks. The Western Mountains ecoregion had the smallest proportion of waterbodies functioning as N~2~O sinks (mean = `r propN2oSinkMeanWM`%, 95% CI: `r propN2oSinkLowerWM` - `r propN2oSinkUpperWM`) and the Northern Plains had the greatest (mean = `r propN2oSinkMeanNPL`%, 95% CI: `r propN2oSinkLowerNPL` - `r propN2oSinkUpperNPL`) **[(Table 1)]**.

@@

# create flextable
flextable(table1_dat) %>%
  set_header_labels('waterbody surface area' = "total waterbody surface area\n(hectares)",
                    'N2O undersaturation extent' = "N2O undersaturation extent\n(% waterbodies)",
                    'N2O emission rate' = "N2O emission rate\n(mg N2O m-2 d-1)",
                    'N2O flux' = "N2O flux\n(metric ton N2O y-1)") %>% # 1 megagram (Mg) = 1 1000 kg = 1 metric ton
  align(j=c(2:6), align = "center", part = "all") %>% # center align columns 2:4
  valign(valign = "top", part = "header") %>% # top align column names
  # caption doesn't show in .Rmd viewer, but renders correctly in .docx
  set_caption(
    caption = as_paragraph( # allows for formatting
      as_chunk("Table 1. Mean and 95% credible interval of N2O undersaturation extent, saturation ratio, emission rate, and flux at the national and ecoregional scales.", 
               props = fp_text_default(bold = TRUE) # here is where we specify formatting (font, size, bold, etc)
      )
    ),
    align_with_table = FALSE) %>% # table is center aligned, ignore that and apply default left align to caption

@@

Despite widespread N~2~O undersaturation in CONUS waterbodies, the mean N~2~O saturation ratio (ratio of measured to equilibrium concentration) was equal to or greater than 1 in seven of the nine ecoregions **[(Table 1, Fig. 2)]**, suggesting that, on average, waterbodies in these ecoregions were supersatured with N~2~O and were therefore functioning as a source of N~2~O. The apparent contradiction between widespread N~2~O undersaturation throughout the CONUS and average dissolved N~2~O concentrations that exceed equilibrium in most ecoregions can be attributed to the distribution of dissolved N~2~O concentrations across the population of CONUS waterbodies **[(Fig. 2)]**. Across the CONUS, and within each of the nine major ecoregions, the distribution of N~2~O saturation ratios had median values that were less than 1, indicating undersaturation, but a long right tail composed of relatively rare waterbodies with levels of N~2~O supersaturation that are sufficiently high to drive the population mean to values close to, or even greater than 1. Thus most waterbodies in the CONUS are functioning as N~2~O sinks during the summer months, but a few function as strong N~2~O sources, particularly at high NO~3~ at water temperature **[Fig. 3]**.

![Figure 2. Distribution of N~2~O saturation values for the population of waterbodies in the nine major ecoregions. Each blue line (n=2000) represents the distribution of values from a single realization of the posterior prediction. Vertical black dashed and solid lines indicate median and mean, respectively. Vertical red line indicates a saturation ratio of 1. x-axis is truncated at 3 to better show the area where 97% of the observations occur.](manuscript_figures/figure2.tiff)

@@

![Figure 3. Conditional effects plot illustrating A) the relationship between N~2~O saturation and nitrate (NO~3~) concentration, dependent upon surface water temperature. The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and \< 4.09, 3) $\geq$ 4.09 and \< 30.20, 4) $\geq$ 30.20 and \< 223.13, and 5) $\geq$ 223.13, and B) relationship between N~2~O saturation and surface water temperature, dependent upon NO~3~ category. Horizontal dashed lines in both panels reflects an N~2~O saturation value of 1.](manuscript_figures/sat_cond_surftemp.tiff)

@@

N~2~O saturation also increased with water temperature **[Fig. 3]**, which is consistent with N~2~O production and consumption occurring through temperature sensitive biological processes, but the relationship was dependent on NO~3~ availability. At low NO~3~ concentrations, temperature had a relatively minor affect on N~2~O saturation, possibly because N~2~O production/consumption was limited by substrate rather than temperature. The effect of temperature became more pronounced when NO~3~ availability was greater, likely reflecting the stimulatory affect of higher temperature on biological N~2~O production. The relative effect of NO~3~ was greater than that of temperature, thus elevated water temperature serves to magnify rates of N~2~O production that are largely dictated by NO~3~ availability.

@@

Persistent patterns in the air-water gas exchange rate across lake size classes may also contribute to greater N~2~O disequilibrium in smaller waterbodies. Air-water gas exchange will vent excess N~2~O from surface waters or facilitate the invasion of air that can replenish depleted N~2~O in undersaturated waterbodies. Regardless of whether a waterbody is superaturated or undersatured in N~2~O, air-water gas exchange will function to move the dissolved N~2~O concentration toward equilibrium. One of the primary drivers of air-water gas exchange in lentic waters is wind speed [@Vachon_etal_2013]. Wind speeds tend to be greater in large lakes with long fetches and lower in small waterbodies, particularly if they are sheltered by shoreline vegetation. Thus, greater air-water gas exchange rates in large lakes may maintain dissolved N~2~O concentrations close to equilibrium, whereas the affect of biogeochemistry on dissolved N~2~O is more pronounced in small waterbodies because low air-water gas exchange limits the rate at which excess N~2~O can be ventilated or depleted N~2~O can be replenished. While previous surveys have documented relationships between lake size and dissolved methane concentration [@Pighini_etal_2018; @Juutinen_etal_2009; @Borges_etal_2011], to the best of our knowledge this is the first study to document a relationship between lake size and $\Delta$ N~2~O.

![Figure 4. A) Relationship between $\Delta$ N~2~O (difference between observed and equilibrium N~2~O concentration) and waterbody surface areas. B,C) relationship between N~2~O saturation and waterbody size for the Coastal Plains and Northern Plains ecoregions. The vertical blue line references a N~2~O saturation value of 1.0. The horizontal red arrows highlight $\Delta$ N~2~O for the smallest lake size class. Mean (points) and 95% credible interview (horizontal bars) are derived from 2000 realizations of the poterior predictive distribution.](manuscript_figures/deltaN2ObySize.tiff)

@@

![Figure 5. Conditional effects plots illustrating the relationship between N~2~O saturation and A) nitrate (NO~3~ availability dependent upon lake size, and B) lake size dependent upon NO~3~ availability. The five NO~3~ categories correspond to concentration ($\mu$g L^-1^) ranges of 1) below detection, 2) $\geq$ 0.55 and \< 4.09, 3) $\geq$ 4.09 and \< 30.20, 4) $\geq$ 30.20 and \< 223.13, and 5) $\geq$ 223.13.](manuscript_figures/sat_cond_area.tiff)

The model indicates that NO~3~ exerts a strong control on N~2~O, but the relationship is complicated by interactions with water temperature and lake surface area. This finding mirrors reports from a survey of farm ponds in Canada where N~2~O was found to be partially controlled by dissolved inorganic nitrogen (DIN) concentration, but the effect of DIN was conditional upon and thermal stratification strength. These complex interactions may explain contrasting literature reports regarding the role of NO~3~ in regulating N~2~O. Several studies have reported positive correlations between NO~3~ and N~2~O in lakes [@Kortelainen_etal_2020; @Yang_etal_2015] while others have found no relationship [@Soued_etal_2016; @Whitfield_etal_2011]. It is possible that the NO~3~ \~ N~2~O relationship wasn't detected in the latter studies because it was confounded by interactions with temperature, lake size, thermal stratification, or other environmental variables.

Interactions between NO~3~, water temperature, and lake size produced a broad range of N~2~O saturation values across the population of CONUS waterbodies, but also enable the identifcation of broad scale patterns. First, small waterbodies with low NO~3~ concentration tend to be undersaturated, whereas small waterbodies with high NO~3~ tend to be highly supersaturated **[Fig. 5]**. Large waterbodies, regardless of NO~3~ status, tend to have N~2~O saturation ratios closer to equilibrium. Therefore intra and inter regional patterns in N~2~O saturation can be well predicted by the joint distributions of waterbody size, NO~3~ availability, and water temperature.

@@

The emission rates reported here span a broad range (`r min.max$min %>% round(.,2)` - `r min.max$max %>% round(.,2)` mg N~2~O m^-2^ d^-1^), which is expected for this large and diverse collection of waterbodies, but are within the range of values reported in the literature for lentic ecosystems [@Yang_etal_2015; @Kortelainen_etal_2020; @Webb_etal_2019]. Despite widespread N~2~O undersaturation in CONUS waterbodies, the national average N~2~O emission rate was positive (mean = `r national_stats$post_med_en2o`, 95% CI: `r national_stats$LCL_en2o` - `r national_stats$UCL_en2o` mg N~2~O m^-2^ d^-1^) **[(Table 1)]**. This contradiction is due to relatively rare waterbodies with very high emission rates that pull the mean (`r national_stats$post_med_en2o` mg N~2~O m^-2^ d^-1^) to a value that far exceeds the median (`r national_stats$post_median_en2o` mg N~2~O m^-2^ d^-1^) **[(SI Fig. 1)]**. At the ecoregional scale, the Northern Appalachians, Northern Plains, and Upper Midwest ecoregions had mean emission rates less than 0. These ecoregions also had the lowest mean NO~3~ availability of all nine ecoregions **[(Table S1)--is this table updated yet?]**.

## Fluxes

Total N~2~O flux from the surface of CONUS waterbodies is a function of the N~2~O emission rate and waterbody surface area. The largest lake size class (\> 50 Ha) contained only `r round(1 - proportionSmallNumber, 2) * 100`% of the total number of lakes in the population, but had `r round(1-proportionSmallSA, 2) * 100`% of total lake surface area. This lake size class had the greatest N~2~O flux, but flux estimates for this lake size class were also the most uncertain **[(Fig. 6)]**. Small uncertainties in emission rates translate to large uncertainties in flux when expressed across the surface area of large waterbodies. Despite widespread N~2~O undersaturation, all but two ecoregions were a net source of N~2~O to the atmosphere **[(Table 1)]**. Differences in N~2~O flux across ecoregions are explained by patterns in N~2~O saturation and waterbody surface **[(Table 1)]**. The Coastal Plains ecoregion had the greatest flux, driven by the high levels of N~2~O saturation and extensive waterbody surface area. By contrast, the Northern Plains had the lowest N~2~O saturation, waterbody surface area, and N~2~O flux.

@@

@@

## Conclusion

Our survey is the first to demonstrate that most lakes and reservoirs in the CONUS function as N~2~O sinks during the summer. Despite widespread N~2~O uptake, CONUS waterbodies function as a net source of N~2~O due to high emissions from from a subset of the population. Dissolved N~2~O in lakes and reservoirs can be predicted from non-linear interactions between water temperature, NO~3~ availability, and waterbody size. NO~3~ availability largely determines whether a waterbody functions as a sink or source of N~2~O, and the magnitude of the source or sink is exaggerated with increasing water temperature. The degree of N~2~O disequilibrium also increases with decreasing waterbody size, possibly because smaller waterbodies have low air-water gas exchange rates.

Although waterbodies in the largest lake size class tend to have N~2~O levels close to equilibrium, they emit more N~2~O than the other size classes due to their expansive surface area. The N~2~O flux from these waterbodies is also more uncertain than the smaller lake size classes because relatively small uncertainties in dissolved N~2~O translate to large uncertainties in flux when propagated across the surface of large waterbodies. While our sample included some of the largest waterbodies in the CONUS, efforts to improve the accuracy of the lentic N~2~O flux estimate should include measurements at the nations largest lakes and reservoirs.

@@

::: {#refs}
:::

@@

## Sampling

Samples were collected from approximately the middle of each waterbody for analysis of algal toxins (microcystins and cylindrospermopsin), benthic macroinvertebrates, phytoplankton, atrazine, water chemistry, chlorophyll‐a, zooplankton, and dissolved N~2~O. Duplicate dissolved N~2~O samples were collected approximately 0.05 m below the water surface by drawing 105 mL of water into a 140 mL syringe equipped with a stopcock and containing 35 mL of air. The syringes were gently shaken for five minutes and the headspace gas transferred to a 30 mL syringe. Finally, the gas sample was transferred into a 12 mL pre-evacuated (\<50 mTorr) glass vial (Labco, Wales, UK) sealed with a PTFE/Silicone stacked on top of a grey chlorobutyl septa. The temperature of the water in the syringe was recorded immediately after the gas was transferred into the vial. Duplicate 20 mL air samples were collected from approximately 2 m above the water surface and transferred to 12 mL glass vials. A vertical profile of water temperature was measured at the same site.
